---
title: "Gender, Education and Vote Choice in the 2024 Presidential Election"
author: "Ana Massoglia"
description: "An analysis of the 2024 ANES Time Series Study"
output:
  distill::distill_article:
    self_contained: false
---
```{r}
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

### Research Question and Hypothesis 

My research question is: How are respondent education level and gender related to voting for Kamala Harris vs. Donald Trump in the 2024 presidential election? For my project, I used the American National Election Studies (ANES) 2024 Time Series Study to examine how social and demographic characteristics relate to public opinion toward presidential candidates.

**My hypothesis is that higher-educated (college+) respondents and women were more likely to vote for Kamala Harris, while lower-educated (high school only) respondents and men were more likely to vote for Donald Trump.** This is plausible because voters often vote in alignment with their identities, so women may feel more positively toward Kamala Harris. Additionally, higher-educated individuals may also be more likely to support Harris because Democratic candidates have increasingly drawn support from college-educated voters, who tend to prioritize issues such as climate policy and social equity. 

**My independent variables of interest are respondent education level and respondent gender.**

**My dependent variable is vote choice for either Harris or Trump.** 

### Why is this important?

Understanding how demographic characteristics like gender and education relate to vote choice is important because it helps explain the political opinions that shape electoral outcomes. The 2024 election was historically significant as the first major-party female presidential nominee faced a former president, making gender dynamics in citizen vote choice particularly relevant. Identifying which groups supported which candidate can inform how parties craft campaign messaging and policy platforms in future elections.

## Data

### Variables and Coding Explanation

In ANES 2024, **education** is measured using a self-reported categorical variable that groups respondents by their highest level of educational attainment (V245015). Because respondents fall into a variety of education categories ranging from high school to post-graduate degrees, there is substantial variation across individuals. 

Respondent **gender** is measured using a self-identified gender variable, which also varies across respondents (V242631). This variation in education and gender enables me to examine whether these characteristics are systematically related to vote choice in the 2024 election. 

My outcome variable is presidential **vote choice** in 2024 (V242067), comparing the Democratic candidate Kamala Harris and the Republican candidate Donald Trump. I recoded this variable so that 1 represents a vote for the Democratic candidate and 0 represents a vote for the Republican candidate, dropping respondents who did not vote for a major-party candidate.

### Criteria for Supporting or Rejecting Hypothesis 

Evidence supporting my hypothesis would be a positive relationship between education level and voting for Kamala Harris, and a positive relationship between being female and voting for the Democratic candidate. Evidence against my hypothesis would include no meaningful relationship or the opposite pattern, such as lower-educated respondents or men being more likely to vote for Harris.

#### Figure 1

```{r}
#| include: false
library(tidyverse)
anes2024 <- read_csv("anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")

```


```{r}
#| include: false
anes2024_small <- anes2024 |>
  select(V242067, V245015, V242631)
```


```{r}
#| include: false
anes2024_clean <- anes2024_small |>
  filter( 
    V242067 %in% c(1, 2),
    V242631 %in% c(1, 2),
    V245015 %in% c(3, 4, 5, 6)
    ) |>
  mutate(
    vote_dem = if_else(V242067 == 1, 1, 0),
    gender = if_else(V242631 == 1, "Male", "Female"),
    education = case_when(
  V245015 == 3  ~ "High school/GED",
  V245015 == 4  ~ "Some college",
  V245015 == 5  ~ "College graduate",
  V245015 == 6  ~ "Master's/PhD"
),
education = factor(education, levels = c(
  "High school/GED", "Some college", "College graduate", "Master's/PhD"
)))
```


```{r}
ggplot(anes2024_clean, aes(x = factor(vote_dem, labels = c("Trump", "Harris")),
                            fill = factor(vote_dem, labels = c("Trump", "Harris")))) +
  geom_bar() +
  scale_fill_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  labs(title = "Vote choice among ANES respondents with 
  complete education and gender data (n = 440)",
       x = "Candidate", y = "Number of respondents",
       caption = "Source: 2024 ANES Time Series Study") +
  theme(legend.position = "none")
```
Figure 1 shows the distribution of presidential vote choice among the 440 ANES respondents included in this analysis. Roughly 56% voted for Harris and 44% for Trump. This sample skews Democratic relative to the actual 2024 election outcome which was a Trump victory. This is a known limitation of the ANES survey design and should be kept in mind when interpreting the results.

## Analysis

#### Figure 2

```{r}
gender_education_summary <- anes2024_clean |>
  group_by(education, gender) |>
  summarize(harris_pct = mean(vote_dem) * 100, .groups = "drop")

ggplot(gender_education_summary, aes(x = education, y = harris_pct, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
  labs(
    title = "Harris vote share by education level and gender",
    caption = "Source: 2024 ANES Time Series Study",
    x = "Education level",
    y = "% voting for Harris",
    fill = "Gender"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = c("Male" = "steelblue", "Female" = "indianred")) 
```
Figure 2 displays Harris vote share broken down by education level and gender. Each bar represents the percentage of respondents in that education and gender group who voted for Harris. The dashed line at 50% indicates the threshold for majority support.



```{r}
#| include: false
fit_1 <- lm(vote_dem ~ education + gender, data = anes2024_clean)

```

#### Table 1

```{r}
fit_1 |>
  broom::tidy() |>
  select(term, estimate, std.error, p.value) |>
  mutate(term = case_when(
    term == "(Intercept)" ~ "Intercept (High school/GED, Female)",
    term == "educationSome college" ~ "Some college",
    term == "educationCollege graduate" ~ "College graduate",
    term == "educationMaster's/PhD" ~ "Master's/PhD",
    term == "genderMale" ~ "Male"
  )) |>
  knitr::kable(digits = 3, col.names = c("Term", "Estimate", "Standard Error", "P-value"))
```
This table displays regression coefficients estimating the relationship between education level, gender and the probability of voting for Harris. The reference category is high school/GED educated female respondents. Estimates represent percentage point differences in the probability of voting for Harris relative to the reference group.

### Additional Analysis

#### Table 2
```{r}
library(modelsummary)

fit_2 <- lm(vote_dem ~ gender, data = anes2024_clean)
fit_3 <- lm(vote_dem ~ education, data = anes2024_clean)
fit_4 <- lm(vote_dem ~ education + gender, data = anes2024_clean)

modelsummary(
  list("Gender Only" = fit_2, 
       "Education Only" = fit_3, 
       "Gender + Education" = fit_4),
  coef_rename = c(
    "genderMale" = "Male",
    "educationSome college" = "Some college",
    "educationCollege graduate" = "College graduate",
    "educationMaster's/PhD" = "Master's/PhD"
  ),
  title = "Table 2: Comparing Models of Harris Vote Choice",
  gof_omit = "IC|Log|Adj|F|RMSE"
)
```


## Discussion of Results 

Figure 2 shows Harris vote share for the 2024 election broken down by education level and gender. Across all four education groups, women voted for Harris at substantially higher rates than men, while the pattern across education levels is less consistent than my hypothesis predicted. 

To examine these relationships, I ran a linear regression of Harris vote choice on education level and gender. The results are shown in Table 1. **The regression confirms that gender is a strong and significant predictor of voting for Harris (p = 0.002).** When holding education constant, male respondents were 14.4 percentage points less likely to vote for Harris than female respondents. I can reject the null hypothesis of no gender effect at the 0.05 significance level. This effect is robust across model specifications, as shown in Table 2, the gender coefficient remains nearly identical (-0.141 vs. -0.144) whether or not education is included in the model.

**In contrast, none of the education coefficients were statistically significant.** Compared to high school/GED respondents, those with some college, a college degree, or a graduate degree do not show meaningfully different rates of Harris support once we account for the p-values. The largest education coefficient is for college graduates (-0.123), but this has a p-value of 0.316, meaning we cannot rule out that this pattern is due to chance. Therefore, I fail to reject the null hypothesis for all education coefficients. This null result also holds in the education-only model shown in Table 2, confirming that the lack of educational effect is not driven by its correlation with gender.

This analysis is cross-sectional and observational, meaning education and gender were not randomly assigned. Therefore, the gender coefficient cannot be interpreted as a causal effect. There are likely confounding variables. For example, party identification is a strong predictor of both gender and vote choice, and omitting it means the gender coefficient may be absorbing some of that variation. Age and race are additional potential confounders. To make a stronger causal claim, one would need either an experimental design or a more complete model that controls for these variables.


## Conclusion

These results partially support my hypothesis. The predicted gender gap confirmed that women were more likely to vote for Harris. However, there is no strong evidence of an educational predictor for Harris support in this sample, which goes against my hypothesis that higher education would predict Harris voting. In fact, high school/GED respondents showed some of the highest support for Harris when looking at my visualization. 

It should be noted that my analysis is limited to the 440 respondents in the ANES sample who voted for a major-party candidate and had complete education and gender data. Respondents with missing gender/education data, those who did not identify as male or female, and those who voted third-party were excluded, which may affect how representative these results are of the 2024 election as a whole.

Additionally, the cross-sectional design means causal claims cannot be made. The most significant threat to inference is potential omitted variable bias. Party identification strongly predicts vote choice and correlates with both gender and education, so the estimated coefficients likely reflect partisan sorting rather than independent demographic effects. If given more time, I would improve this analysis by controlling for party identification, race and age.

##### AI Disclosure
AI was used in the specific instance of inquiring about how to remove values that were not needed in my analysis. The suggestion was using gof_omit = "IC|Log|Adj|F|RMSE" to clean up the second table. 

## Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
