---
title: "PS270 Final Project"
author: "Ana Massoglia"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project proposal

My research question is: How are household income and respondent gender related to voting for Kamala Harris vs. Donald Trump the 2024 presidential election? For my final project, I will use the American National Election Studies (ANES) 2024 Time Series Study to examine how social and economic characteristics relate to public opinion toward presidential candidates. My hypothesis is that lower-income respondents and women were more likely to vote for Kamala Harris, while higher-income respondents and men were more likely to vote for the Donald Trump. This is plausible because voters often vote in alignment with their identities, so women may feel more positively toward Kamala Harris. Additionally, lower-income individuals may also be more likely to support Kamala Harris because Democratic policies are often more closely associated with economic redistribution and expanded social programs. My explanatory variables of interest are household income and respondent gender. In ANES 2024, household income is measured using an ordered categorical variable that groups respondents into income brackets (V241555). Because respondents fall into a variety of income categories, there is substantial variation across individuals. Respondent gender is measured using a self-identified gender variable, which also varies across respondents (V242631). This variation in income and gender enables me to examine whether these characteristics are systematically related to vote choice in the 2024 election. My outcome variable is presidential vote choice in 2024 (V242067z), comparing the Democratic candidate Kamala Harris and Republican candidate Donald Trump. I will recode this variable so that 1 represents a vote for the Democratic candidate and 0 represents a vote for the Republican candidate, dropping respondents who did not vote for a major-party candidate. Evidence supporting my hypothesis would be a negative relationship between income and voting for the Kamala Harris, and a positive relationship between being female and voting for the Democratic candidate. Evidence against my hypothesis would include no meaningful relationship or the opposite pattern, such as higher-income respondents or men being more likely to vote for Harris.

```{r}
library(tidyverse)
anes2024 <- read_csv("anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")
head(anes2024)
```
```{r}
"V242067z" %in% names(anes2024)
"V241555" %in% names(anes2024)
"V242631" %in% names(anes2024)
```
```{r}
anes2024_small <- anes2024 |>
  select(V242067z, V241555, V242631)
head(anes2024_small)
```

