---
title: "PS270 Final Project"
author: "Ana Massoglia"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project proposal

My research question is: How are respondent education level and gender related to voting for Kamala Harris vs. Donald Trump in the 2024 presidential election? For my final project, I will use the American National Election Studies (ANES) 2024 Time Series Study to examine how social and demographic characteristics relate to public opinion toward presidential candidates. My hypothesis is that higher-educated (college+) respondents and women were more likely to vote for Kamala Harris, while lower-educated (high school only) respondents and men were more likely to vote for Donald Trump. This is plausible because voters often vote in alignment with their identities, so women may feel more positively toward Kamala Harris. Additionally, higher-educated individuals may also be more likely to support Harris because Democratic candidates have increasingly drawn support from college-educated voters, who tend to prioritize issues such as climate policy and social equity. My explanatory variables of interest are respondent education level and respondent gender. In ANES 2024, education is measured uses a self-reported categorical variable that groups respondents by their highest level of educational attainment (V245015). Because respondents fall into a variety of education categories ranging from high school to post-graduate degrees, there is substantial variation across individuals. Respondent gender is measured using a self-identified gender variable, which also varies across respondents (V242631). This variation in education and gender enables me to examine whether these characteristics are systematically related to vote choice in the 2024 election. My outcome variable is presidential vote choice in 2024 (V242067), comparing the Democratic candidate Kamala Harris and the Republican candidate Donald Trump. I will recode this variable so that 1 represents a vote for the Democratic candidate and 0 represents a vote for the Republican candidate, dropping respondents who did not vote for a major-party candidate. Evidence supporting my hypothesis would be a positive relationship between education level and voting for Kamala Harris, and a positive relationship between being female and voting for the Democratic candidate. Evidence against my hypothesis would include no meaningful relationship or the opposite pattern, such as lower-educated respondents or men being more likely to vote for Harris.

```{r}
library(tidyverse)
anes2024 <- read_csv("anes_timeseries_2024_csv_20250808/anes_timeseries_2024_csv_20250808.csv")
head(anes2024)
```
```{r}
"V242067" %in% names(anes2024)
"V245015" %in% names(anes2024)
"V242631" %in% names(anes2024)
```
```{r}
anes2024_small <- anes2024 |>
  select(V242067, V245015, V242631)
head(anes2024_small)
```
```{r}
anes2024_clean <- anes2024_small |>
  filter( 
    V242067 %in% c(1, 2),
    V242631 %in% c(1, 2),
    V245015 %in% c(3, 4, 5, 6)
    ) |>
  mutate(
    vote_dem = if_else(V242067 == 1, 1, 0),
    gender = if_else(V242631 == 1, "Male", "Female"),
    education = case_when(
  V245015 == 3  ~ "High school/GED",
  V245015 == 4  ~ "Some college",
  V245015 == 5  ~ "College graduate",
  V245015 == 6  ~ "Master's/PhD"
),
education = factor(education, levels = c(
  "High school/GED", "Some college", "College graduate", "Master's/PhD"
)))
anes2024_clean
```

```{r}
gender_education_summary <- anes2024_clean |>
  group_by(education, gender) |>
  summarize(harris_pct = mean(vote_dem) * 100, .groups = "drop")

ggplot(gender_education_summary, aes(x = education, y = harris_pct, fill = gender)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray40") +
  labs(
    title = "Harris vote share by education level and gender",
    caption = "Source: 2024 ANES Time Series Study",
    x = "Education level",
    y = "% voting for Harris",
    fill = "Gender"
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_fill_manual(values = c("Male" = "steelblue", "Female" = "indianred")) 
```

